The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-04-01 Paper.Rmd
edits for Discussion

<!--8) A para about what is available in the copula and VineCopula packages. Maybe make it a broader para about emthods, and state in there that we developed some new surrogate methods that may be broadly useful for researches trying to incorporate copula approaches.-->
Copula approaches are
well developed, statistically [@joe2014_dependence; @nelsen2006_copula], and 
computer implementations exist (e.g., the `copula` and 
`VineCopula` packages in R). Ecologists need only to become aware of these tools
and understand them conceptually to start applying them broadly. 
We created a variety of interrelated randomization procedures
(Appendices \ref{SM-surrog_test}, \ref{SM-Surrog_multi}, \ref{SM-sect:copsurrognd}) 
that build upon existing methods and may be especially useful for ecological applications.

@@

<!--Maybe a para on how likely one is to get copula structure via the Moran mechanism - considers
work on copula structure in environmental variables. Maybe mention possible future work: 
study copula structure of environmental variables (such as temperature and precip) that are known 
to be important in ecology, including how such copula structure may be changing and whether it 
has a geography of the kind studied in Walter et al.-->
<!--***DAN: Shya, please read or reread all the papers we have that say anything about 
copula structure of dependence between an environmental variable measured in different places, 
and then write this paragraph.-->
One may expect our Moran-like mechanism for non-N copula structure and tail dependence to 
operate commonly, because Moran effects are likely ubiquitous [@Liebhold2004], and 
non-N and tail-dependent copula structures are common in environmental variables.
<!--***DAN: Shya, this topic sentence may have to be modified if you find non-N and tail-dependent
copula structures in environmental variables measured in different places are not so common,
or have not been studied much. Butif you find some papers, and they show non-N structure, you 
can just summarize their results here.-->
<!--***DAN: Idea to work in if it fits: if large weather events (e.g., large rain events) 
are also widespread, that should lead to right-tail dependence in, e.g., rainfall.-->

<!--Maybe a para on how likely one is to get copula structure via the Liebig mechanism - considers
what people have written on nonlinear effects and empirical evidence for Liebig's law and so on-->
There are also substantial reasons to expect that our second mechanism operates commonly. 
The second mechanism is the one that involves asymmetric or nonlinear influences of 
environmental variables. Liebig's law of the minimum and the idea of the "limiting 
nutrient" for organism or population growth are fundamental and dominant paradigms in ecology.
Furthermore, innumerable studies document nonlinear influences of environmental
variables on important ecological quantities at all levels of organization, e.g., the 
influence of temperature on: 
individual metabolism and autotroph growth rates 
[@GarciaCarreras2018]; dispersal [@Oconnor2007]; population growth 
[@Savage2004]; and microbial and ecosystem methane flux [@YvonDurocher2014].
<!--***DAN: you may want to combine the above para with the previous one in the end, using an
intro something like "There are good reasons to suspect that the two mechanisms we suggested
operate extremely commonly in real ecosystems"-->

<!--4) We have only given two mechanisms, but there are likely to be others, and discoverig
additional mechanisms of copula structure is certain worthy of future work. Maybe I can think
of some ideas and make a paragraph.
5) If Van gives us some ideas, they may fit in the Discussion here.
6) Maybe give Holder a bunch of draws from a bivariate dist with standard normals marginals but with 
some tail-dependent copula structure and ask him to simulate evolution down a mammal or bird tree and
see what you get. Then give him the other tail dependence, and no tail dependence (bivariate standard
normal). Maybe this will lead to more results, and also a Discussion para on why evolutionary steps 
might be tail dependent. A result that tail dependent steps lead to tail dependent character dists
in the end would replace one question (why are there tail-dependent characters) with another (why are
there tail dependent steps). Of course it could also be the case that steps are not tail dependent and 
the tail dependence of characters arises in another way. Would also lead in to alternative thinking in 
evolutionary biology - Mark could write a para on why it would matter for evolutionary biol if there are 
tail dependent steps.
-->

@@

<!--a para on stability of the total biomass time series in a community, and skewness - sets up Shya's later paper as future work-->

@@

<!--Brief conclusion of some kind-->

>>>>

<!--8) A para about what is available in the copula and VineCopula packages. Maybe make it a broader para about emthods, and state in there that we developed some new surrogate methods that may be broadly useful for researches trying to incorporate copula approaches.-->
Copula approaches are
well developed, statistically [@joe2014_dependence; @nelsen2006_copula], and 
computer implementations exist (e.g., the `copula` and 
`VineCopula` packages in R). Ecologists need only become aware of these tools
and understand them conceptually to start applying them broadly. 
We created a variety of interrelated randomization procedures
(Appendices \ref{SM-surrog_test}, \ref{SM-Surrog_multi}, \ref{SM-sect:copsurrognd}) 
that build upon existing methods and may be especially useful for ecological applications.

@@

<!--Maybe a para on how likely one is to get copula structure via the Moran mechanism - considers
work on copula structure in environmental variables. Maybe mention possible future work: 
study copula structure of environmental variables (such as temperature and precip) that are known 
to be important in ecology, including how such copula structure may be changing and whether it 
has a geography of the kind studied in Walter et al.-->
One may expect our Moran-like mechanism for non-N copula structure and tail dependence to 
operate commonly, because Moran effects are common [@Liebhold2004], and 
non-N and tail-dependent copula structures have often been found in prior analyses of 
environmental variables. If large meteorological events are also spatially widespread, then
one would expect environmental variables associated with these events to take extreme
values simultaneously across large spatial areas, producing tail dependence in measurements
made at different locations. Non-extreme values, not associated with extreme events, may
be more associated with local weather phenomena, and therefore 
less prone to being correlated across large areas.
<!--***Shya: Done but please check the last line of this para : 
@Goswami2018 this paper tests model with 3 different copula : Frank, Gaussian, Plackett, as far as I know all are symmetric, 
so non-N copula may be best fit but in this case I could not understand how tha tail-dep. thing would fit. 
***DAN: that's OK, the purpose here is to give background on what has been done, it does not all have to pit perfectly. And also, the topic sentence just says non-N, which that is.--> 
Using a bivariate mixed model approach with copulas, 
@Serinaldi2008 examined the spatial dependence 
of rainfall observations (7 years with 30 minute time resolution) 
within a rain gauge network in Central Italy. Gumbel or Student 2-copulas 
were possible candidates for modelling bivariate dependence at several time scales, and 
neither of these is an N copula. @liu2016probabilistic used a copula 
based approach in assessing the probability of vegetation related drought across China. 
<!--***DAN: Shya, what do you mean by "vegetation related drought"? Is this a kind of drought
that is somehow caused by the vegetation? That's what it sounds like, and it sounds like what
they studied was the probability that such a drought occurred. If that's not what you mean, please
rephrase.-->
In a majority of their cases, they found 
non-N copula structure in the dependence between a vegatation index 
and climatic variables (e.g., precipitation and temperature). 
Another long-term study (1950–2014) in the Loess Plateau of China [@She2018]
showed that a Gumbel copula (which has upper tail dependence) could effectively model 
the spatial dependence of drought variables 
(duration and severity) in the study area. Bivariate copula analysis was also used in forecasting the 
co-occurrence of extreme events (flood or drought) 
over the North Sikkim Himalayas, India, using spatial data sets at different 
time scales [@Goswami2018].

<!--Maybe a para on how likely one is to get copula structure via the Liebig mechanism - considers
what people have written on nonlinear effects and empirical evidence for Liebig's law and so on-->
There are also good reasons to expect that our second mechanism operates commonly. 
The second mechanism is the one that involves asymmetric or nonlinear influences of 
environmental variables. Liebig's law of the minimum and the idea of a "limiting 
nutrient" of organism or population growth are fundamental and dominant paradigms in ecology.
Furthermore, innumerable studies document nonlinear influences of environmental
variables on important ecological quantities at all levels of organization, e.g., the 
influence of temperature on: 
individual metabolism and autotroph growth rates 
[@GarciaCarreras2018]; dispersal [@Oconnor2007]; population growth 
[@Savage2004]; and microbial and ecosystem methane flux [@YvonDurocher2014].
Environmental variables influence not only ecological populations, but also physiological,
community, and ecosystem functioning measurements. So our two mechanism may well operate
to produce spatial copula structure in a wide variety of ecological measurements.

<!--4) We have only given two mechanisms, but there are likely to be others, and discoverig
additional mechanisms of copula structure is certain worthy of future work. Maybe I can think
of some ideas and make a paragraph.
5) If Van gives us some ideas, they may fit in the Discussion here.
6) Maybe give Holder a bunch of draws from a bivariate dist with standard normals marginals but with 
some tail-dependent copula structure and ask him to simulate evolution down a mammal or bird tree and
see what you get. Then give him the other tail dependence, and no tail dependence (bivariate standard
normal). Maybe this will lead to more results, and also a Discussion para on why evolutionary steps 
might be tail dependent. A result that tail dependent steps lead to tail dependent character dists
in the end would replace one question (why are there tail-dependent characters) with another (why are
there tail dependent steps). Of course it could also be the case that steps are not tail dependent and 
the tail dependence of characters arises in another way. Would also lead in to alternative thinking in 
evolutionary biology - Mark could write a para on why it would matter for evolutionary biol if there are 
tail dependent steps.
-->
Other mechanisms that we have not considered likely also operate to produce copula structure
in ecological datasets.
For instance, we detected non-N structure and tail dependence in our 
metabolic rate-versus-body mass datasets,
but the mechanisms discussed above do not apply directly in a macroevolutionary context.
We hope by enumerating two potential mechanisms we help inspire additional theoretical and
empirical research on the
potentially numerous mechanisms of copula structure that may apply in diverse ecological datasets.
<!--***DAN: beef this up if/when we get inputs from Mark and Van-->

@@

<!--a para on stability of the total biomass time series in a community, and skewness - sets up Shya's later paper as future work-->
Our results showed that the skewness of the 
spatial average of local time series from different locations is influenced 
by their tail dependence. But the logic of this result should also apply to any collection of time
series, whether associated with different locations in space or not, that are summed or averaged. 
Another potential application is time series of all species from a single community, e.g., all
plants in a quadrat or survey site (REF) or all herbivorous plankton in a lake (REF). A huge 
literature has focused on synchrony versus compensatory dynamics between such time series (REFS), and 
the influence of such interspecific relationships on the variability, through time, of total-community
properties such as total biomass. Typically, variability through time of total-community
properties is measured with the coefficient of variation, but the skewness of such a time series
may also be of interest: it can characterize the extent to
which community properties show "spikiness," i.e., uncharacteristically extreme values. 
Therefore, future work on copula structure of intraspecific relationships in communities and its
influence on community properties seems likely to us to be a potentially very exciting avenue.

@@
